### 2019-07-08 thesis-template/msc-thesis.tex
update

\begin{itemize}
	\item a light transport model (propagation scheme) following basic but crucial physical principles,
	\item visualization of the propagation results in $2D$ polar profiles and scalar fields,
	\item application of this model for tensor field visualization interpreting tensors as light transmission properties,
	\item a FTLE (Finite-time Lyapunov exponents)-related approach called light transport gradient (LTG) for visualizing key structures, namely LCS (Lagrangian coherent structures) in $2D$ second-order tensor fields, and
	\item application of our approach to both synthetic and real data involving brain and heart datasets.
\end{itemize}
%Dieses Kapitel gibt einen ?berblick ?ber die Arbeit. Gerade der
%Abschnitt zur Motivation soll allgemein verst?ndlich geschrieben
%werden. Die Einleitung sollte auch wichtige Referenzen enthalten.

@@

The main purpose of visualization in general, is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor - medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue, or furthermore in simulated or measured stress tensor fields from solid and fluid mechanics. In these named domains of science the problem arises, that tensor fields can be generated or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e., mostly 2D/3D-grids, commonly consisting of second-order tensors (cf.~Cauchy stress tensor). In general, these grids span a spatial domain with sample positions indicated by vertices or cell centers in the dual grid, depending on the mode of interpretation. Tensor fields feature a sampling grid, consistent of tensors, which is spaced uniformly in the commonly used regular grid. A Cartesian grid, which we will utilize in this context, is a grid normalized to a certain unit length, as a special case of a regular grid. That is, each cell is formed as a unit cube (square) in relation to the normalizing unit length. We aim to use global illumination methods to model light transport, directed by tensor fields, within such type of grids, which we consider as light propagation volumes (LPVs) \cite{dachsbacher}. Congruently, the most basic objective of this work is to design a simple light propagation scheme following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, eventually to visualize tensor fields. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (e.g., $r(\omega)=1$) by the user which we sample (discretize) as a primary step. This way, the approach will operate as a light transport visualization technique for light distributions given in polar coordinates, provided the user set an initial profile into the grid. Furthermore, this propagation scheme should either be used to visualize the light transport paths, defined by the anisotropy characteristics of tensor fields and tensor magnitudes of a given tensor field in a 2D grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a $2D$ grid. However, the tensor field needs to be preprocessed somehow to generate a footprint map of transmission profiles defined by the tensor field's anisotropy characteristics, which we will handle through PCA yielding a principal component ellipsoid. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of  directive effect on the light transport controlled by the tensor's anisotropy characteristics. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. This will yield an approach, capable of computing approximated final light distributions for a given light source configuration inside of a 2D lattice. As this process requires a lot of user-interaction and is not to be concerned as generalizing to all possible scenarios, we aim to do something more elaborate on the basis of the light propagation scheme. We place light sources in form of Dirac pulses at any position and pointing in any direction represented in the grid, which represents a flow map, which we will analyze for its gradient to reveal the topolgical structure of the tensor field. In the sense of automization and optimization, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to grasp by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (i.e., forming a saddle) local features, which frequently represent key structures in nature, such as moraines in glaciers or vortices in flow fields. Commonly, in vector field visualization, LCS separate regions of different flow behavior, but in our case they separate regions, or in other words subdomains, of varying directional distributions (manifesting in diverging/converging tensor field lines leading to compressions, tensions, roations or varying tensor magnitude causing volume changes) represented by tensors, as we use our tensor-controlled light transport visualization (propagation scheme) as a basis or framework to compute the FTLE field. Consequently, we propose another method to compute an FTLE-like field on tensor fields. The FTLE field responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other conceptually, because of bidirectional stress directions. Since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases dependent on our initial light source direction. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide new tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain abstracted representations and the interpretation of meaningful results evaluating the proposed techniques for possible uses. The designed approach could, e.g., be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations. A classical example for tensor field visualization is plotting hyperstreamlines following major/minor eigenvectors, as depicted in Fig.~\ref{streamlines}, which features a structurally unstable point (purple), a wedge (red) and a trisector (green).

@@

There has been extensive recent work on symmetric tensors but in comparison little on asymmetric ones. Since we are not able to cover the whole scope in this work, we will focus on the most relevant works on global illumination and tensor field visualization. We will talk about the publications from Hlawatsch et al. \cite{hlawatsch} and Zirr et al. \cite{zirr} in particular, since they constitute thematically close related works. But first, we will cover related works on global illumination methods, as these form a basis or prerequisite for our work. Please note, that they do not focus on tensor field visualization at all. Remember, that we will inventively employ global illumination methods for tensor field visualization as an innovative step and that we want to cover previous works on our base approach, as well.

\section{Global Illumination Methods}
Since we use a specific category of global illumination methods as a basis, we decide to only discuss lattice based methods which work, corresponding to their name, on a grid. Remark, that Global Illumination methods classically focus on realistic indirect lighting. Thus, they aim on producing a most realistic image, what in our case turns out to be a most informative image in correspondence. The most early work in this field is propably the Discrete Ordinates Method (DOM) by Chandrasekhar \cite{Chandrasekhar}, which is used specifically for computing radiative transfer in participating media (for volume rendering). The radiative transfer equation is discretized spatially and in the angular domain. The radiance distributions are represented in a grid and light is distributed and exchanged between neighboring volume elements which reduces the overall effort to local and independent computations. Furthermore, there has been more recent work on describing the light propagation as a diffusion process within a simple photon transport model, based on the Lattice-Boltzmann method \cite{geist}. DOMs were recently revisited by Fattal \cite{fattal}. This work addresses artifacts emerging from spatial discretization, which are called \enquote{ray effects}, and from successive interpolation called \enquote{light smearing}. These methods are considered to be too slow to \enquote{run at interactive framerates} by Dachsbacher et al. \cite{dachsbacher}, though \enquote{efficient and potentially highly parallel}. Our work is inspired by Lattice-Boltzmann methods (volume grid radiosity) and the lattice-based discrete ordinate methods concerning the light propagation scheme.

@@

Zheng \& and Pang \cite{pang&zheng} et al. proposed a texture-based visualization approach called HyperLIC, which extends the concept of LIC to symmetric tensor fields by using an anisotropic 2D-filter kernel oriented along the major or minor eigenvectors. The concept of following the major/minor eigenvector along tensor field lines was first invented by Vilanova et al. \cite{vilanova}. Tensorlines \cite{weinstein} introduce a kind of artificial inertia (running average), making them more resistent to noise. The notion of tensor field topology (critical, degenerate points, separatrices) and the concept of hyperstreamlines was first introduced by Delmarcelle and Hesselink \cite{delmarcelle&hesselink}. The topology in $3D$ tensor fields was first analyzed by Zheng et al. \cite{zheng}. The placement of tensor field lines as hyperstreamlines for glyph packing was recently improved by Spencer et al. \cite{spencer}. A good overview of seeding strategies concerning glyphs is given by McLoughlin et al. \cite{mcloughlin}. Feng et al. \cite{feng} used Voronoi tessellation for placing the glyphs. Superquadric glyphs have been proposed by Schultz and Kindlmann \cite{schultz&kindlmann} for general (non-positive definite) tensors. De Leeuw and van Wijk \cite{deleeuw&vanwijk} visualize the partial derivative gradient, the Jacobian of the tensor field, to sense the local properties of the field. A symmetric tensor can also be visualized by box \cite{makris}, ellipsoid \cite{pierpaoli}, composite \cite{westin2} or superquadric \cite{kindlmann} glyphs. There is also a set of scalar measures available in symmetric tensor field visualization such as: mean diffusivity, fractional anisotropy \cite{basser} and anisotropy coefficients \cite{westin}. These coefficients can also be visualized by direct volume rendering \cite{kindlmann2}. Hlawatsch et al. \cite{hlawatsch} obtain a coordinate (flow) map from tensor field lines following major/minor eigenvectors (fibre trajectories) using the maximum eigenvalue of the Cauchy-Green deformation tensor to compute the FTLE-field in homogeneous regions. In a similar manner, we aim to use Global Illumination methods propagating light, directed by tensor fields, to obtain a final light distribution (flow map) through successive light propagation cycles. The light distributions are redirected by the tensor field employing transmission profiles obtained from ellipsoid glyphs. DT-MRI diffusion tensor imaging can help to reveal the functional \cite{denis} or topological \cite{wakana} structure of the brain and muscles by analysing celebral and muscle tissue tractography, which yields a tensor field representing the anisotropy characteristcs of the brain's white matter. An overview about DT-MRI visualization techniques is given by \cite{vilanova}. Tricoche et al. \cite{xavier} use invariant manifolds to extract topological features from tensor fields. Other feature extraction-based approaches extract general features as: edges \cite{hancock}, tensor shape\/ orientation \cite{gordon}, interfaces \cite{donnel} and creases \cite{tricochet}.

@@

Zheng and Pang \cite{pang&zheng} proposed the concept of dual eigenvectors and Zhang et al. \cite{zhang} extended it to pseudo-eigenvectors and introduced the eigenvector/eigenvalue manifolds to visualize eigenvectors in the complex domain. Laramee et al. \cite{laramee} focused on the efficient implementation and visualization of these structures and provided an interactive visualization system for asymmetric tensors applicable in fluid and solid dynamics. The concept of tensor magnitude has been introduced by \cite{laramee} for means of physical interpretation. They also proposed an efficient glyph and hyperstreamline hybrid approach, which made dynamic interaction in real-time in $2D$ tensor fields feasible. Palacios et al. \cite{palacios} extracted isosurfaces of tensor magnitude, mode and isotropy index.

Last, we will discuss our work in relation to other publications. While these previously mentioned works use classical approaches like tensor field lines and glyphs, we will also introduce the concept of Lagrangian coherent structures into our framework contribution to tensor field analysis. LCS have been defined as time-dependent analog of separatrices \cite{haller} which are concerned to be robust under noisy conditions \cite{haller2}, which we expect from our technique as well by implication. Hlawatsch et al. \cite{hlawatsch} obtain similar results (a FTLE-like field on tensor fields), but obtain them by sampling of fiber trajectories instead of global illumination light distributions, as it is the case for this work. Our work shares the same basic idea, deriving and computing a FTLE-like field for tensor field visualization generalizing FTLE from vectors to tensors, yet using a light propagation scheme and hence using whole distributions of trajectories incorporating polar profiles. Concerning the tensor field visualization method (LTG), our approach is based on FTLE, and principal component analysis (PCA). We will detect LCS with an FTLE-derived method considering differences in final light transport distributions through computing the gradient of the resulting flow maps. The FTLE-related approach named ``light transport gradient'', to be designed, also states a generalization of the light transport visualization method FTPD proposed by Zirr et al. \cite{zirr}, since it allows a similar measure to be computed through defining a geometrical scene and setting the transmission (defined by the tensor field) to an isotropic and constant value of $100\%$. In this operating mode, the approach operates as a light transport visualization technique neglecting any tensor field transmission profiles and is thus capable of displaying bounding separatrices of a light transport scenario perturbated by obstacles and mirrors \cite{zirr}. 
%Typischerweise im letzten Abschnitt dieses Kapitels wird dann auf
%verwandte Arbeiten eingegangen. Entsprechende Arbeiten sind geeignet
%zu zitieren. Beispiel: Die wurde erstmalig in den Arbeiten von Spitz
%und Gertz \cite{Spitz2016a} gezeigt \ldots Details dazu werden in
%dem Buch von Newman zu Netzwerken \cite{Newman2010} erl?utert \ldots.

@@

\section{Tensor Fields}
\label{sec:tensors}
Now what is a tensor? A tensor is an $o$-dimensional generalization of a matrix as illustrated in Table \ref{tensor}.
\begin{table}[!b]
\centering
\caption{Tensor Shapes}
 \begin{tabular}{c|c|c|c|c|c|c}
\textbf{order} & 0 & 1 & 2 & 3 & ... & o \\ 
\hline 
\textbf{shape} & scalar & vector & matrix & ``3D matrix'' & ... & order-$o$ matrix\\ 
\end{tabular}
\label{tensor}
 \end{table} 
In addition to that, a tensor has as many indices as its order $o$ and their run length is as long as its embedded dimension $n$ in space (the dimension of the row and column vectors), which is equal to the rank of the tensor for full rank. Or in other words the number of column elements $n$ for a $n{\times}n$-matrix. A tensor is considered non-degenerate if all eigenvalues $\lambda_i>\lambda_{i+1}$ manifesting in significant anisotropy. Tensors follow certain transformation rules which are defined for covariant and contravariant tensors, which indicate the change characteristics under a basis change. Contravariant tensors do not change under a change of basis, whereas covariant tensors are not invariant w.r.t. basis changes. Tensors are most commonly obtained and defined from mathematical transformation equations. Hence, any order-$o$ number array could be a tensor, but the definition holds only if the transformation rules apply to these $o$-th order number arrays (in a sense of mathematical/formal interpretation). This is mostly true for component indexed order-$o$ number arrays following matrix multiplication and transformation rules found in physics and math manifesting in calculation rules like matrix multiplication. Thus, a tensor in general is a order-$o$ number array arranged in meaningful (transformation) order, to put it simple. In fact, tensors themselves are interpreted as transformations which reflect an incoming direction vector (measuring, e.g., the stress in that direction) to an outgoing (stress) vector. To be a bit more concrete, tensors represent stress distributions in solids and fluids describing strain and flow features, as these entities are not simple numbers or vectors. They occur in form of a tensors, and thus as entities or quantities in physics. That is, at each point (typically a Euclidean space or manifold) in space (indexed by the grid) there is a whole distribution of directions, e.g., stresses (elastic and viscous), which needs to be characterized and represented by a single tensor each. Collectively, defined at each point in space, these tensors form a tensor field.
\paragraph{stress tensor}
\begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.6\textwidth]
    {img/stress_tensor.png}
  }
  \caption{Cauchy stress Tensor}
  \captionsetup{font={footnotesize,bf,it}}
  \caption*{source: \url{https://en.wikipedia.org/wiki/Cauchy_stress_tensor\#/media/File:Components_stress_tensor_cartesian.svg}}
  \label{cauchy}
\end{figure}

@@

\section{Finite-Time Lyapunov Exponent}
\label{sec:ftle}
The Finite-Time Lyapunov Exponent is a measure, developed to analyze time-dependent dynamical systems for their separation abilities concerning massless tracer particles. We imagine a flow field, i.e., a vector field which in turn represents the effective amount or concentration shifted into a particular direction at each point in space. It could be induced by e.g. concentration or height gradients. This flow field is considered a time-dependent dynamical system, since it usually holds a time-component $t$ and thus maps each initial position an aged (altered) position: $\mathbf{\upvarphi}(\mathbf{x}) : \mathbf{x}(t_0) \longmapsto \mathbf{x}(t_0+T)$. We can now imagine placing two close by particles in distance $\delta$ into a flow field and run a time dependent simulation. Remark that this corresponds to the Lagrangian view in vector fields, which travels with the particles (instead of taking the position of a resting observer called Eulerian view).

@@

In Fig. \ref{ftle_sep}, we observe particle trajectories diverging w.r.t. the central axis ending up in distance $\Delta$. The FTLE measure for this single example (point in space) is then defined as:

@@

\begin{figure}[!t]
\centering
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/corner-TFL.png}
	a)``corner''
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/spiral-TFL.png}
	b)``spiral''
    \label{b)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/drain_old.png}
	c)``V''
    \label{b)}
  \end{minipage}
\caption{Several test fields}
\label{test-fields}
\end{figure}

\begin{figure}[!t]
\centering
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/drain(alt)-TFL.png}
	a)
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/tensorfieldlines.png}
	b)
    \label{b)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/inverse-TFL.png}
	c)
    \label{b)}
  \end{minipage}
\caption{Several test fields: a)``corner'', ``spiral'', ``V''}
\label{test_fields2}
\end{figure}

>>>>

\begin{itemize}
	\item a light transport model (propagation scheme) following basic but crucial physical principles,
	\item visualization of the propagation results in 2D polar profiles and scalar fields,
	\item application of this model for tensor field visualization interpreting tensors as light transmission properties,
	\item a FTLE (Finite-time Lyapunov exponents)-related approach called light transport gradient (LTG) for visualizing key structures, namely LCS (Lagrangian coherent structures) in 2D second-order tensor fields, and
	\item application of our approach to both synthetic and real data involving brain and heart datasets.
\end{itemize}
%Dieses Kapitel gibt einen ?berblick ?ber die Arbeit. Gerade der
%Abschnitt zur Motivation soll allgemein verst?ndlich geschrieben
%werden. Die Einleitung sollte auch wichtige Referenzen enthalten.

@@

The main purpose of visualization in general, is to reduce or encode some set of numbers in some readable, explorable and more intuitive representation. This implies, e.g., clustering, structuring or projection techniques. When it comes to tensor fields, this comes in handy in, e.g., medical sciences such as in DT-MRI (diffusion tensor - medical resonance imaging) \cite{hlawatsch}, when measuring water molecule diffusion characteristics within tissue, or furthermore in simulated or measured stress tensor fields from solid and fluid mechanics. In these named domains of science the problem arises, that tensor fields can be generated or measured with physical/mathematical background, but not be regarded intuitively. Necessity begets ingenuity as always, which lets visualization methods and systems emerge in this context. Concurrently, the problem setting of this work encomprises the visualization of tensor fields, i.e.~, mostly 2D/3D-grids, commonly consisting of second-order tensors (cf.~Cauchy stress tensor). In general, these grids span a spatial domain with sample positions indicated by vertices or cell centers in the dual grid, depending on the mode of interpretation. Tensor fields feature a sampling grid, consistent of tensors, which is spaced uniformly in the commonly used regular grid. A Cartesian grid, which we will utilize in this context, is a grid normalized to a certain unit length, as a special case of a regular grid. That is, each cell is formed as a unit cube (square) in relation to the normalizing unit length. We aim to use global illumination methods to model light transport, directed by tensor fields, within such type of grids, which we consider as light propagation volumes (LPVs) \cite{dachsbacher}. Congruently, the most basic objective of this work is to design a simple light propagation scheme following the most basic but crucial physical principles and laws, like energy conservation and propagation attenuation, eventually to visualize tensor fields. Of course, the propagation scheme will need some kind of stimulus which can be defined through symbolic functions (e.g., $r(\omega)=1$) by the user which we sample (discretize) as a primary step. This way, the approach will operate as a light transport visualization technique for light distributions given in polar coordinates, provided the user set an initial profile into the grid. Furthermore, this propagation scheme should either be used to visualize the light transport paths, defined by the anisotropy characteristics of tensor fields and tensor magnitudes of a given tensor field in a 2D grid. Or, as a secondary purpose, be used to visualize the light transport in a defined, given geometrical scene in a 2D grid. However, the tensor field needs to be preprocessed somehow to generate a footprint map of transmission profiles defined by the tensor field's anisotropy characteristics, which we will handle through PCA yielding a principal component ellipsoid. For this purpose, we will let the transmission profiles act on our simple light propagation scheme to have a kind of  directive effect on the light transport controlled by the tensor's anisotropy characteristics. The transmission profiles should modulate the outgoing intensities in a sense of redirection or more appropriately, redistribution. This will yield an approach, capable of computing approximated final light distributions for a given light source configuration inside of a 2D lattice. As this process requires a lot of user-interaction and is not to be concerned as generalizing to all possible scenarios, we aim to do something more elaborate on the basis of the light propagation scheme. We place light sources in form of Dirac pulses at any position and pointing in any direction represented in the grid, which represents a flow map, which we will analyze for its gradient to reveal the topolgical structure of the tensor field. In the sense of automization and optimization, another consequent aim is to implement an approach borrowed from vector-field visualization to reveal structures, which are hard to grasp by ordinary methods. We construct an FTLE-related approach in tensor field visualization (both symmetric and asymmetric) for detecting ridges representing LCS (Lagrangian Coherent Structures). These structures can be classified (identified) as repelling, attracting and saddling (i.e.~, forming a saddle) local features, which frequently represent key structures in nature, such as moraines in glaciers or vortices in flow fields. Commonly, in vector field visualization, LCS separate regions of different flow behavior, but in our case they separate regions, or in other words subdomains, of varying directional distributions (manifesting in diverging/converging tensor field lines leading to compressions, tensions, roations or varying tensor magnitude causing volume changes) represented by tensors, as we use our tensor-controlled light transport visualization (propagation scheme) as a basis or framework to compute the FTLE field. Consequently, we propose another method to compute an FTLE-like field on tensor fields. The FTLE field responds most where tensor field lines / hyperstreamlines converge or diverge, whereas one is the same as the other conceptually, because of bidirectional stress directions. Since we effectively measure the finite-time Lyapunov exponent (FTLE) with both forward and backward integration time (stimulus directions, in our case), we will detect both cases dependent on our initial light source direction. Both invented methods, will give new insights into the topology and anisotropy of tensor fields and provide new tools for the structural analysis. The greater goal of this work is therefore the encoding of a tensor field into certain abstracted representations and the interpretation of meaningful results evaluating the proposed techniques for possible uses. The designed approach could, e.g., be applied in medicinal sciences, engineering and physics to reveal structures which are hard to grasp through interpreting simpler and sometimes confusing glyph and tensor field line visualizations. A classical example for tensor field visualization is plotting hyperstreamlines following major/minor eigenvectors, as depicted in Fig.~\ref{streamlines}, which features a structurally unstable point (purple), a wedge (red) and a trisector (green).

@@

There has been extensive recent work on symmetric tensors but in comparison little on asymmetric ones. Since we are not able to cover the whole scope in this work, we will focus on the most relevant works on global illumination and tensor field visualization. We will talk about the publications from Hlawatsch et al.~\cite{hlawatsch} and Zirr et al.~\cite{zirr} in particular, since they constitute thematically close related works. But first, we will cover related works on global illumination methods, as these form a basis or prerequisite for our work. Please note, that they do not focus on tensor field visualization at all. Remember, that we will inventively employ global illumination methods for tensor field visualization as an innovative step and that we want to cover previous works on our base approach, as well.

\section{Global Illumination Methods}
Since we use a specific category of global illumination methods as a basis, we decide to only discuss lattice based methods which work, corresponding to their name, on a grid. Remark, that Global Illumination methods classically focus on realistic indirect lighting. Thus, they aim on producing a most realistic image, what in our case turns out to be a most informative image in correspondence. The most early work in this field is propably the Discrete Ordinates Method (DOM) by Chandrasekhar \cite{Chandrasekhar}, which is used specifically for computing radiative transfer in participating media (for volume rendering). The radiative transfer equation is discretized spatially and in the angular domain. The radiance distributions are represented in a grid and light is distributed and exchanged between neighboring volume elements which reduces the overall effort to local and independent computations. Furthermore, there has been more recent work on describing the light propagation as a diffusion process within a simple photon transport model, based on the Lattice-Boltzmann method \cite{geist}. DOMs were recently revisited by Fattal \cite{fattal}. This work addresses artifacts emerging from spatial discretization, which are called \enquote{ray effects}, and from successive interpolation called \enquote{light smearing}. These methods are considered to be too slow to \enquote{run at interactive framerates} by Dachsbacher et al.~\cite{dachsbacher}, though \enquote{efficient and potentially highly parallel}. Our work is inspired by Lattice-Boltzmann methods (volume grid radiosity) and the lattice-based discrete ordinate methods concerning the light propagation scheme.

@@

Zheng \& and Pang \cite{pang&zheng} et al.~proposed a texture-based visualization approach called HyperLIC, which extends the concept of LIC to symmetric tensor fields by using an anisotropic 2D-filter kernel oriented along the major or minor eigenvectors. The concept of following the major/minor eigenvector along tensor field lines was first invented by Vilanova et al.~\cite{vilanova}. Tensorlines \cite{weinstein} introduce a kind of artificial inertia (running average), making them more resistent to noise. The notion of tensor field topology (critical, degenerate points, separatrices) and the concept of hyperstreamlines was first introduced by Delmarcelle and Hesselink \cite{delmarcelle&hesselink}. The topology in 3D tensor fields was first analyzed by Zheng et al.~\cite{zheng}. The placement of tensor field lines as hyperstreamlines for glyph packing was recently improved by Spencer et al.~\cite{spencer}. A good overview of seeding strategies concerning glyphs is given by McLoughlin et al.~\cite{mcloughlin}. Feng et al.~\cite{feng} used Voronoi tessellation for placing the glyphs. Superquadric glyphs have been proposed by Schultz and Kindlmann \cite{schultz&kindlmann} for general (non-positive definite) tensors. De Leeuw and van Wijk \cite{deleeuw&vanwijk} visualize the partial derivative gradient, the Jacobian of the tensor field, to sense the local properties of the field. A symmetric tensor can also be visualized by box \cite{makris}, ellipsoid \cite{pierpaoli}, composite \cite{westin2} or superquadric \cite{kindlmann} glyphs. There is also a set of scalar measures available in symmetric tensor field visualization such as: mean diffusivity, fractional anisotropy \cite{basser} and anisotropy coefficients \cite{westin}. These coefficients can also be visualized by direct volume rendering \cite{kindlmann2}. Hlawatsch et al.~\cite{hlawatsch} obtain a coordinate (flow) map from tensor field lines following major/minor eigenvectors (fibre trajectories) using the maximum eigenvalue of the Cauchy-Green deformation tensor to compute the FTLE-field in homogeneous regions. In a similar manner, we aim to use Global Illumination methods propagating light, directed by tensor fields, to obtain a final light distribution (flow map) through successive light propagation cycles. The light distributions are redirected by the tensor field employing transmission profiles obtained from ellipsoid glyphs. DT-MRI diffusion tensor imaging can help to reveal the functional \cite{denis} or topological \cite{wakana} structure of the brain and muscles by analysing celebral and muscle tissue tractography, which yields a tensor field representing the anisotropy characteristcs of the brain's white matter. An overview about DT-MRI visualization techniques is given by \cite{vilanova}. Tricoche et al.~\cite{xavier} use invariant manifolds to extract topological features from tensor fields. Other feature extraction-based approaches extract general features as: edges \cite{hancock}, tensor shape\/ orientation \cite{gordon}, interfaces \cite{donnel} and creases \cite{tricochet}.

@@

Zheng and Pang \cite{pang&zheng} proposed the concept of dual eigenvectors and Zhang et al.~\cite{zhang} extended it to pseudo-eigenvectors and introduced the eigenvector/eigenvalue manifolds to visualize eigenvectors in the complex domain. Laramee et al.~\cite{laramee} focused on the efficient implementation and visualization of these structures and provided an interactive visualization system for asymmetric tensors applicable in fluid and solid dynamics. The concept of tensor magnitude has been introduced by \cite{laramee} for means of physical interpretation. They also proposed an efficient glyph and hyperstreamline hybrid approach, which made dynamic interaction in real-time in 2D tensor fields feasible. Palacios et al.~\cite{palacios} extracted isosurfaces of tensor magnitude, mode and isotropy index.

Last, we will discuss our work in relation to other publications. While these previously mentioned works use classical approaches like tensor field lines and glyphs, we will also introduce the concept of Lagrangian coherent structures into our framework contribution to tensor field analysis. LCS have been defined as time-dependent analog of separatrices \cite{haller} which are concerned to be robust under noisy conditions \cite{haller2}, which we expect from our technique as well by implication. Hlawatsch et al.~\cite{hlawatsch} obtain similar results (a FTLE-like field on tensor fields), but obtain them by sampling of fiber trajectories instead of global illumination light distributions, as it is the case for this work. Our work shares the same basic idea, deriving and computing a FTLE-like field for tensor field visualization generalizing FTLE from vectors to tensors, yet using a light propagation scheme and hence using whole distributions of trajectories incorporating polar profiles. Concerning the tensor field visualization method (LTG), our approach is based on FTLE, and principal component analysis (PCA). We will detect LCS with an FTLE-derived method considering differences in final light transport distributions through computing the gradient of the resulting flow maps. The FTLE-related approach named ``light transport gradient'', to be designed, also states a generalization of the light transport visualization method FTPD proposed by Zirr et al.~\cite{zirr}, since it allows a similar measure to be computed through defining a geometrical scene and setting the transmission (defined by the tensor field) to an isotropic and constant value of $100\%$. In this operating mode, the approach operates as a light transport visualization technique neglecting any tensor field transmission profiles and is thus capable of displaying bounding separatrices of a light transport scenario perturbated by obstacles and mirrors \cite{zirr}. 
%Typischerweise im letzten Abschnitt dieses Kapitels wird dann auf
%verwandte Arbeiten eingegangen. Entsprechende Arbeiten sind geeignet
%zu zitieren. Beispiel: Die wurde erstmalig in den Arbeiten von Spitz
%und Gertz \cite{Spitz2016a} gezeigt \ldots Details dazu werden in
%dem Buch von Newman zu Netzwerken \cite{Newman2010} erl?utert \ldots.

@@

\section{Tensor Fields}
\label{sec:tensors}
Now what is a tensor? A tensor is an $o$-dimensional generalization of a matrix as illustrated in Table \ref{tensor}.
\begin{table}[!b]
\centering
\caption{Tensor Shapes}
 \begin{tabular}{c|c|c|c|c|c|c}
\textbf{order} & 0 & 1 & 2 & 3 & ... & o \\ 
\hline 
\textbf{shape} & scalar & vector & matrix & ``3D matrix'' & ... & order-$o$ matrix\\ 
\end{tabular}
\label{tensor}
 \end{table} 
In addition to that, a tensor has as many indices as its order $o$ and their run length is as long as its embedded dimension $n$ in space (the dimension of the row and column vectors), which is equal to the rank of the tensor for full rank. Or in other words the number of column elements $n$ for a $n{\times}n$-matrix. A tensor is considered non-degenerate if all eigenvalues $\lambda_i>\lambda_{i+1}$ manifesting in significant anisotropy. Tensors follow certain transformation rules which are defined for covariant and contravariant tensors, which indicate the change characteristics under a basis change. Contravariant tensors do not change under a change of basis, whereas covariant tensors are not invariant w.r.t.~basis changes. Tensors are most commonly obtained and defined from mathematical transformation equations. Hence, any order-$o$ number array could be a tensor, but the definition holds only if the transformation rules apply to these $o$-th order number arrays (in a sense of mathematical/formal interpretation). This is mostly true for component indexed order-$o$ number arrays following matrix multiplication and transformation rules found in physics and math manifesting in calculation rules like matrix multiplication. Thus, a tensor in general is a order-$o$ number array arranged in meaningful (transformation) order, to put it simple. In fact, tensors themselves are interpreted as transformations which reflect an incoming direction vector (measuring, e.g., the stress in that direction) to an outgoing (stress) vector. To be a bit more concrete, tensors represent stress distributions in solids and fluids describing strain and flow features, as these entities are not simple numbers or vectors. They occur in form of a tensors, and thus as entities or quantities in physics. That is, at each point (typically a Euclidean space or manifold) in space (indexed by the grid) there is a whole distribution of directions, e.g., stresses (elastic and viscous), which needs to be characterized and represented by a single tensor each. Collectively, defined at each point in space, these tensors form a tensor field.
\paragraph{stress tensor}
\begin{figure}[!t]
  \centering
 {
    \includegraphics[width=0.6\textwidth]
    {img/stress_tensor.png}
  }
  \caption{Cauchy stress Tensor}
  \captionsetup{font={footnotesize,bf,it}}
  \caption*{source: \url{https://en.wikipedia.org/wiki/Cauchy_stress_tensor\#/media/File:Components_stress_tensor_cartesian.svg}}
  \label{cauchy}
\end{figure}

@@

\section{Finite-Time Lyapunov Exponent}
\label{sec:ftle}
The Finite-Time Lyapunov Exponent is a measure, developed to analyze time-dependent dynamical systems for their separation abilities concerning massless tracer particles. We imagine a flow field, i.e.~, a vector field which in turn represents the effective amount or concentration shifted into a particular direction at each point in space. It could be induced by e.g. concentration or height gradients. This flow field is considered a time-dependent dynamical system, since it usually holds a time-component $t$ and thus maps each initial position an aged (altered) position: $\mathbf{\upvarphi}(\mathbf{x}) : \mathbf{x}(t_0) \longmapsto \mathbf{x}(t_0+T)$. We can now imagine placing two close by particles in distance $\delta$ into a flow field and run a time dependent simulation. Remark that this corresponds to the Lagrangian view in vector fields, which travels with the particles (instead of taking the position of a resting observer called Eulerian view).

@@

In Fig.~\ref{ftle_sep}, we observe particle trajectories diverging w.r.t.~the central axis ending up in distance $\Delta$. The FTLE measure for this single example (point in space) is then defined as:

@@

\begin{figure}[!t]
\centering
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/corner-TFL.png}
	a)``Corner''
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/spiral-TFL.png}
	b)``Spiral''
    \label{b)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=\textwidth]{img/drain_old.png}
	c)``V''
    \label{b)}
  \end{minipage}
\caption{Several test fields}
\label{test-fields}
\end{figure}

\begin{figure}[!t]
\centering
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/drain(alt)-TFL.png}
	a)
    \label{a)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/tensorfieldlines.png}
	b)
    \label{b)}
  \end{minipage}
  \begin{minipage}{0.3\textwidth}
      \centering
    \includegraphics[height=\textwidth]{img/inverse-TFL.png}
	c)
    \label{b)}
  \end{minipage}
\caption{Several test fields: a)``Drain'', ``Rings'', ``Inverse''}
\label{test_fields2}
\end{figure}
